---
title: "Assignment 3 - Part1"
author: "Shreya"
date: "25 October 2016"
output: html_document
---

K-means clustering algorithm

```{r}

install.packages("ggplot2") 

library(dplyr)
library(tidyr)
library(ggplot2)

```

upload the file "Class_Motivation.csv" 
```{r}

K1 <- read.table("E:/TC/4050/GITHUB/assignment3/Class_Motivation.csv", header = TRUE, sep = ",")

```

This file contains the self-reported motivation scores for a class over five weeks. We are going to look for patterns in motivation over this time and sort people into clusters based on those patterns.

But before we do that, we will need to manipulate the data frame into a structure that can be analyzed by our clustering algorithm.

The algorithm will treat each row as a value belonging to a person, so we need to remove the id variable.

```{r}

K2 <- dplyr::select(K1, 2:6)

K3 <- na.omit(K2) 
#This command create a data frame with only those people with no missing values. It "omits" all rows with missing values, also known as a "listwise deletion". EG - It runs down the list deleting rows as it goes.

```

Another pre-processing step used in K-means is to standardize the values so that they have the same range. We do this because we want to treat each week as equally important - if we do not standardise then the week with the largest range will have the greatest impact on which clusters are formed. We standardise the values by using the "scales()" command.

```{r}

K3 <- scale(K3)

```


Now we will run the K-means clustering algorithm we talked about in class. 
1) The algorithm starts by randomly choosing some starting values 
2) Associates all observations near to those values with them
3) Calculates the mean of those clusters of values
4) Selects the observation closest to the mean of the cluster
5) Re-associates all observations closest to this observation
6) Continues this process until the clusters are no longer changing

Notice that in this case we have 5 variables and in class we only had 2. It is impossible to vizualise this process with 5 variables.

Also, we need to choose the number of clusters we think are in the data. We will start with 2.

```{r}

fit <- kmeans(K3, 2) 

#We have created an object called "fit" that contains all the details of our clustering including which observations belong to each cluster.

#We can access the list of clusters by typing "fit$cluster", the top row corresponds to the original order the rows were in. Notice we have deleted some rows.

fit$cluster

#We can also attach these clusters to the original dataframe by using the "data.frame" command to create a new data frame called K4.

K4 <- data.frame(K3, fit$cluster)

#Have a look at the K4 dataframe. Lets change the names of the variables to make it more convenient with the names() command.

names(K4) <- c("1", "2", "3", "4", "5", "cluster") #c() stands for concatonate and it creates a vector of anything, in this case a vector of names.


```

Now we need to visualize the clusters we have created. To do so we want to play with the structure of our data. What would be most useful would be if we could visualize average motivation by cluster, by week. To do this we will need to convert our data from wide to long format. Remember your old friends tidyr and dplyr!

First lets use tidyr to convert from wide to long format.
```{r}

K5 <- tidyr::gather(K4, "week", "motivation", 1:5)

```

Now lets use dplyr to average our motivation values by week and by cluster.

```{r}

K6 <- K5 %>% group_by(week, cluster)
K6 <- summarise(K6, avg = mean(motivation))

```

Now it's time to do some visualization! We are going to start using the ggplot2 package, a very popular visualization package in R. It is based on the "Grammar of Graphics" a theory of how visualizations work best. 


```{r}

K6$week <- as.numeric(K6$week)

K6$cluster <- as.factor(K6$cluster)

```

Now we can plot our line plot using the ggplot command, "ggplot()".

- The first argument in a ggplot is the dataframe we are using: K6
- Next is what is called an aesthetic (aes), the aesthetic tells ggplot which variables to use and how to use them. Here we are using the variables "week" and "avg" on the x and y axes and we are going color these variables using the "cluster" variable
- Then we are going to tell ggplot which type of plot we want to use by specifiying a "geom()", in this case a line plot: geom_line()
- Finally we are going to clean up our axes labels: xlab("Week") & ylab("Average Motivation")

```{r}

ggplot(K6, aes(week, avg, colour = cluster)) + geom_line() + xlab("Week") + ylab("Average Motivation")

#Rplot1-1

```

What patterns do you see in the plot?

It would be useful to determine how many people are in each cluster. We can do this easily with dplyr.

```{r}
K7 <- dplyr::count(K4, cluster)
```
______________________________________________________________________________________


Look at the number of people in each cluster, now repeat this process for 3 rather than 2 clusters. 

```{r}

K8 <- na.omit(K2)
K8 <- scale(K8)

fit1 <- kmeans(K8, 3) 
fit1$cluster

K9 <- data.frame(K8, fit1$cluster)

names(K9) <- c("1", "2", "3", "4", "5", "cluster")

K10 <- tidyr::gather(K9, "week", "motivation", 1:5)

K11 <- K10 %>% group_by(week, cluster)


K11 <- summarise(K11, avg = mean(motivation))


K11$week <- as.numeric(K11$week)
K11$cluster <- as.factor(K11$cluster)

library(ggplot2)
ggplot(K11, aes(week, avg, colour = cluster)) + geom_line() + xlab("Week") + ylab("Average Motivation")
#Rplot1-2

K11 <- dplyr::count(K9, cluster)
```

Which cluster grouping do you think is more informative? Write your answer below:

According to me, the Rplot2 makes more sense or is more informative. SInce, there are 3 clusters we can differntiate them more and on comparing we see 
- the group 1 at the top, in red is sinking on the level of motivation over the weeks
- the group 3 in blue almost inthe middle, started with a negative motivation. Its trying hard, fluctuating but motivation level seems to improve.
- the group 2 in green, at the bottom is not motivated to a normal level that should be there, but is rather negative. 

Hence, group 2 need more attention and external motivation may be. Group 3 is trying good.
Group 1 needs to pull its strings, because it started good but is sinking negatively over the weeks.


Done..!!