# Assignment 3 - Social Network Analysis

## Part I
Start by installing the "igraph" package. Once you have installed igraph, load the package.

Now upload the data file "comment-data.csv" as a data frame called "D1". Each row represents a comment from one student to another so the first line shows that student "28" commented on the comment of student "21". It also shows the gender of both students and the students' main elective field of study ("major"").

```{r}
#install.packages("igraph")
library(igraph)

D1 <- read.csv("comment-data.csv", header = TRUE)
```

Before you proceed, you will need to change the data type of the student id variable. Since it is a number R will automatically think it is an integer and code it as such (look at the list of variables by clicking on the data frame arrow in the Data pane. Here you will see the letters "int"" next to the stid variable, that stands for integer). However, in this case we are treating the variable as a category, there is no numeric meaning in the variable. So we need to change the format to be a category, what R calls a "factor". We can do this with the following code:

```{r}
D1$comment.to <- as.factor(D1$comment.to)
D1$comment.from <- as.factor(D1$comment.from)

table(D1$comment.from)
table(D1$comment.to)
```

igraph requires data to be in a particular structure. There are several structures that it can use but we will be using a combination of an "edge list" and a "vertex list" in this assignment. As you might imagine the edge list contains a list of all the relationships between students and any characteristics of those edges that we might be interested in. There are two essential variables in the edge list a "from" variable and a "to" variable that descibe the relationships between vertices. While the vertex list contains all the characteristics of those vertices, in our case gender and major.

So let's convert our data into an edge list!

First we will isolate the variables that are of interest: comment.from and comment.to

```{r}
library(dplyr)

D2 <- select(D1, comment.to, comment.from) #select() chooses the columns
```

Since our data represnts every time a student makes a comment there are multiple rows when the same student comments more than once on another student's video. We want to collapse these into a single row, with a variable that shows how many times a student-student pair appears.

```{r}

EDGE <- count(D2, comment.to, comment.from)

names(EDGE) <- c("to", "from", "count")

```

EDGE is your edge list. Now we need to make the vertex list, a list of all the students and their characteristics in our network. Because there are some students who only recieve comments and do not give any we will need to combine the comment.from and comment.to variables to produce a complete list.

```{r}
#First we will separate the commenters from our commentees
V.FROM <- select(D1, comment.from, from.gender, from.major)

#Now we will separate the commentees from our commenters
V.TO <- select(D1, comment.to, to.gender, to.major)

#Make sure that the from and to data frames have the same variables names
names(V.FROM) <- c("id", "gender.from", "major.from")
names(V.TO) <- c("id", "gender.to", "major.to")

#Make sure that the id variable in both dataframes has the same number of levels
lvls <- sort(union(levels(V.FROM$id), levels(V.TO$id)))

VERTEX <- full_join(mutate(V.FROM, id=factor(id, levels=lvls)),
    mutate(V.TO, id=factor(id, levels=lvls)), by = "id")

# Create a copy of gender from/to, and keep them as characters. This will help label the graph, while also 
# allowing us to use them as factors (Charles' code below) to code colors based on major
VERTEX$major.char <- ifelse(is.na(VERTEX$major.from) == TRUE, VERTEX$major.to, VERTEX$major.from)

# Fill in missing gender and major values - ifelse() will convert factors to numerical values so convert to character
VERTEX$gender.from <- ifelse(is.na(VERTEX$gender.from) == TRUE, as.factor(as.character(VERTEX$gender.to)), as.factor(as.character(VERTEX$gender.from)))

VERTEX$major.from <- ifelse(is.na(VERTEX$major.from) == TRUE, as.factor(as.character(VERTEX$major.to)), as.factor(as.character(VERTEX$major.from)))

#Remove redundant gender and major variables
VERTEX <- select(VERTEX, id, gender.from, major.from, major.char)

#rename variables
names(VERTEX) <- c("id", "gender", "major", "major.char")

#Remove all the repeats so that we just have a list of each student and their characteristics
VERTEX <- unique(VERTEX)

```

Now we have both a Vertex and Edge list it is time to plot our graph!

```{r}
#Load the igraph package

library(igraph)

#First we will make an object that contains the graph information using our two dataframes EDGE and VERTEX. Notice that we have made "directed = TRUE" - our graph is directed since comments are being given from one student to another.

g <- graph.data.frame(EDGE, directed=TRUE, vertices=VERTEX)

#Now we can plot our graph using the force directed graphing technique - our old friend Fruchertman-Reingold!

plot(g,layout=layout.fruchterman.reingold)

#There are many ways to change the attributes of the graph to represent different characteristics of the newtork. For example, we can color the nodes according to gender.

plot(g,layout=layout.fruchterman.reingold, vertex.color=VERTEX$gender)

#We can change the thickness of the edge according to the number of times a particular student has sent another student a comment.

plot(g,layout=layout.fruchterman.reingold, vertex.color=VERTEX$gender, edge.width=EDGE$count)

````

## Part II

In Part II your task is to [look up](http://igraph.org/r/) in the igraph documentation and modify the graph above so that:

* Ensure that sizing allows for an unobstructed view of the network features (For example, the arrow size is smaller)
* The vertices are colored according to major
* The vertices are sized according to the number of comments they have recieved


```{r}
#?igraph.plotting

# Define Graph
  g <- graph.data.frame(EDGE, directed=TRUE, vertices=VERTEX)

# Generate colors for nodes
  colors <- c("yellow4", "tomato", "gold", "cadetblue")

# Vertex Char.
  V(g)$color <- colors[V(g)$major] # vertex colors based on majors
  V(g)$size <- E(g)$count*6 #vertex node size comes from a var in the Edge frame
  V(g)$frame.color="black" #surrounding line of vertex, same color as edge
  V(g)$label.color <- "black" #default is blue, change to black
  V(g)$label.cex=.5 # label size for labels
  V(g)$label.dist=1
  V(g)$label=NA # turn it off for now
  V(g)$legend=V(g)$major

# Edge Char.
  E(g)$width <- .3 #Edge width - make it thinner
  E(g)$arrow.size <- .2 # make it smaller to unobstruct view
  E(g)$edge.color <- "gray80"

# For Legend: Find out number/Char equivalency. 
# There is certainly a be way to do it, but anyway...
  plot(g) 
  title("Student Comment Network",cex.main=1,col.main="black")
  legend(x=1.5, y=-1.1, c("M1", "M2", "M3", "M4"), pch=21, pt.bg=colors, pt.cex=2, cex=.8, bty="n", ncol=1)




```


## Part III

Now practice with data from our class. This data is real class data directly exported from Qualtrics and you will need to wrangle it into shape before you can work with it. Import it into R as a data frame and look at it carefully to identify problems.

Please create a **person-network** with the data set hudk4050-classes.csv. To create this network you will need to create a person-class matrix using the tidyr functions and then create a person-person matrix using `t( )`. You will then need to plot a matrix rather than a to/from data frame using igraph.


### STEP1: DATA WRANGLING
```{r}

#install.packages("stringr")
#install.packages("foreach")
  library(stringr)
  library(dplyr)
  library(tidyr)
  library(tidyverse)
  library(foreach)
  library(igraph)


# Import data
  DF <- read.csv("hudk4050-classes.csv", header = TRUE)
 
# Rename variables
  names(DF) <-  unlist(DF[1,])
  names(DF) <- sub(" ", ".", names(DF))

#remove first TWO rows
  DF <- DF[-c(1, 2), ] 

# Create numeric ID 
  DF$ids <- 1:47
  DF$ids<- as.factor(DF$ids)
  DF$interest <- DF$`Which.of these topics is most interesting to you?`

#Make data frame with all classes and numeric ID. 
#DF2 <- data.frame(DF$ids, DF[,grep("Class", colnames(DF))] ) #also works, but select(, contains()) is better
  DF2 <- select(DF, ids, contains("Class"))
  
# All uppercase
  DF2 <- data.frame(lapply(DF2, function(v) {
    if (is.character(v)) return(toupper(v))
    else return(v)
  }))
  
# Get rid of space between dept and cn 
  DF2 <- data.frame(lapply(DF2, function(v) {
    if (is.character(v)) str_replace(v, " ", "")
    else return(v)
  }))

# From Wide to Long
  DF3 <- DF2 %>% gather("Class.n", "class", -ids)
  #DF2 <- select(DF2, class, ids)
  table(DF3$class)
  
# Get rid of missing entries 
  DF3$class <- str_replace(DF3$class, " ", "")
  DF3 <- DF3 %>% filter(str_detect(DF3$class, "[0-9]"))

# From long to wide
  DF3$dummy <- rep(1, 163)
  
  #diff# DF3 <- DF2 %>% spread(ids, dummy) # one entry per class 
  DF4 <- DF3 %>% spread(class, dummy) # one entry per student
  
# Collapse to the student level
  DF4 <- select(DF4, -Class.n)
  names(DF4) = paste0("class_", names(DF4))
  DF4 <- DF4 %>% rename(ID=class_ids)
  
  DF4[is.na(DF4)] = 0
  
  DF5 <- DF4 %>% group_by(ID) %>% summarise_if(is.numeric, max)
  #diff# DF4 <- DF3 %>% group_by(class) %>% summarise_if(is.numeric, max)
  #diff# names(DF4) = paste0("std_", names(DF4))
  #diff# DF4 <- DF4 %>% rename(class = std_class) 
  
# Data level: STD 
# There are 52 course variables in the data 
# If two students are in the same course, a connection should be created
  
  
  

```


### STEP2: TRY MATRIX FROM DF5 (STUDENT LEVEL)
```{r}

## Aha! This does not work because R works columnwise! ugh
EDGE = matrix(c(0,0),ncol = 2)
for(c in 2:53){                          # 52 classes in wide format
  for(id1 in 1:47){                      # 47 students
  for(id2 in 1:47){                      # 47 students
    if(DF5[id1,c] == DF5[id2,c] & DF5[id1,c]==1 & id1!=id2){   # if 2 dif std have a "1" for the same class: connect
      EDGE = rbind(EDGE, matrix(c(DF5[id1,1],DF5[id2,1]), ncol = 2))
    }
  }
  }
}

```
  
### STEP3: TRY MATRIX FROM DF3 (STUDENT/COURSE LEVEL)
  
```{r}

## Let's go back to DF3 and try columnwise
### Question: how to remove duplicates? example: 1/2 & 2/1
  
DF3$c <- as.factor(DF3$class)
  
EDGE = matrix(c(0,0,0),ncol = 3)
  
  for(o1 in 1:163){                      # total obs from long dataset
  for(o2 in 1:163){                      # total obs from long dataset
    if(DF3[o1,5] == DF3[o2,5] & o1!=o2) {   # if 2 std are in the same class
      EDGE = rbind(EDGE, matrix(c(DF3[o1,1],DF3[o2,1], DF3[o1, 5]), ncol = 3))
    }
  }
  }

EDGE = EDGE[-1,]
colnames(EDGE) = c("std1", "std2", "class")

```



### STEP4: EDGE AND VERTEX FROM MATRIX

```{r}

  # EDGE: To/From IDs
  # Vertex: All IDs 
  
  VERTEX <- select(DF, First.Name, ids, interest)
  VERTEX <- VERTEX %>% rename(ID=ids)
  VERTEX$interest <- as.factor(as.character(VERTEX$interest))

  EDGE <- as.data.frame(EDGE)
  EDGE <- count(EDGE, std1, std2)
  
```


### STEP5: GRAPH


```{r}

  g2 <- graph.data.frame(EDGE, directed=F, vertices=VERTEX$ID)
  #g2 <- simplify(g2, remove.multiple=TRUE)
  #is.simple(g2)
  
# Generate colors for nodes
  colors <- c("yellow4", "tomato", "gold", "cadetblue", "gray80")

# Vertex Char.
  V(g2)$color <- colors[VERTEX$interest] 
  V(g2)$size <- 10
  V(g2)$frame.color="black" #surrounding line of vertex, same color as edge
  V(g2)$label.color <- "black" #default is blue, change to black
  V(g2)$label.cex=.5 # label size for labels
  V(g2)$label.dist=0

# Edge Char.
  E(g2)$width <- EDGE$n*.2 #Edge width - based on the n of connections between std
  E(g2)$arrow.size <- .2 # make it smaller to unobstruct view
  E(g2)$edge.color <- "gray80"

# Plot
  
  #why isn't simplify command above working when I knit this? 
  # It DOES work when I run the code chunk
  
  plot(g2, layout=layout.fruchterman.reingold, remove.multiple=TRUE) 
  title("Student Course-taking Network",cex.main=1,col.main="black")

```



Once you have done this, also [look up](http://igraph.org/r/) how to generate the following network metrics:

* Betweeness centrality and dregree centrality. **Who is the most central person in the network according to these two metrics? Write a sentence or two that describes your interpretation of these metrics**

```{r}

## ON MY GRAPH (G2)
#Degree centrality - this was supposed to indicate the individual with most connections. unclear in this network.
sort(degree(g2), decreasing=TRUE)

#betweeness - individual 27 is the shortest path to other individuals. this could be considered the most "influencial" person in the network.
sort(betweenness(g2), decreasing=TRUE)



```


* Color the nodes according to interest. Are there any clusters of interest that correspond to clusters in the network? Write a sentence or two describing your interpetation.

```{r}

# nodes already colored according to interest

plot(g2, layout=layout_nicely)
  title("Student Course-taking Network",cex.main=1,col.main="black")

```


# Graph after Code workout :-) 
```{r}

# Could work more on this - will come back to it later!

# Transfer IDS to ROWNAMES - rownames are already the num IDS, but I liked this command
rownames(DF5) <- DF5$ID
# Remove 4050, OK
DF6 <- select(DF5, -class_HUDK4050)
# Remove ID column
DF6 <- select(DF6, -ID)
# Matrix - classes as rows, without names
CW <- as.matrix(DF6)
#Craete person-person
CW <- CW %*% t(CW)
#Graphing 
g3 <- graph.adjacency(CW, mode="undirected", diag=FALSE)


# Vertex Char.
  V(g3)$size <- 10
  V(g3)$frame.color="black" 
  V(g3)$label.color <- "black" 
  V(g3)$label.cex=.5 # label size for labels
  V(g3)$label.dist=0


plot (g3)



```


### To Submit Your Assignment

Please submit your assignment by first "knitting" your RMarkdown document into an html file and then comit, push and pull request both the RMarkdown file and the html file.
