# Assignment 3 - Social Network Analysis

## Part I
Start by installing the "igraph" package. Once you have installed igraph, load the package.

Now upload the data file "comment-data.csv" as a data frame called "D1". Each row represents a comment from one student to another so the first line shows that student "28" commented on the comment of student "21". It also shows the gender of both students and the students' main elective field of study ("major"").

```{r}
D1 <- read.csv("comment-data.csv", header = TRUE)
```

Before you proceed, you will need to change the data type of the student id variable. Since it is a number R will automatically think it is an integer and code it as such (look at the list of variables by clicking on the data frame arrow in the Data pane. Here you will see the letters "int"" next to the stid variable, that stands for integer). However, in this case we are treating the variable as a category, there is no numeric meaning in the variable. So we need to change the format to be a category, what R calls a "factor". We can do this with the following code:

```{r}
D1$comment.to <- as.factor(D1$comment.to)
D1$comment.from <- as.factor(D1$comment.from)
```

igraph requires data to be in a particular structure. There are several structures that it can use but we will be using a combination of an "edge list" and a "vertex list" in this assignment. As you might imagine the edge list contains a list of all the relationships between students and any characteristics of those edges that we might be interested in. There are two essential variables in the edge list a "from" variable and a "to" variable that descibe the relationships between vertices. While the vertex list contains all the characteristics of those vertices, in our case gender and major.

So let's convert our data into an edge list!

First we will isolate the variables that are of interest: comment.from and comment.to

```{r}
library(dplyr)

D2 <- select(D1, comment.to, comment.from) #select() chooses the columns
```

Since our data represnts every time a student makes a comment there are multiple rows when the same student comments more than once on another student's video. We want to collapse these into a single row, with a variable that shows how many times a student-student pair appears.

```{r}

EDGE <- count(D2, comment.to, comment.from)

names(EDGE) <- c( "to", "from","count")

```

EDGE is your edge list. Now we need to make the vertex list, a list of all the students and their characteristics in our network. Because there are some students who only recieve comments and do not give any we will need to combine the comment.from and comment.to variables to produce a complete list.

```{r}
#First we will separate the commenters from our commentees
V.FROM <- select(D1, comment.from, from.gender, from.major)

#Now we will separate the commentees from our commenters
V.TO <- select(D1, comment.to, to.gender, to.major)

#Make sure that the from and to data frames have the same variables names
names(V.FROM) <- c("id", "gender.from", "major.from")
names(V.TO) <- c("id", "gender.to", "major.to")


#Make sure that the id variable in both dataframes has the same number of levels
lvls <- sort(union(levels(V.FROM$id), levels(V.TO$id)))

#So this union will work on a on a vector. It's kind of like a join, but just for a vector. And then I'm going to use the levels command.That extracts the levels from each variable. So we're going to take out that this one is 25(i.e. v.FROM$id) and this one that has 22 and then union is going to join those two things together for us. And then because

VERTEX <- full_join(mutate(V.FROM, id=factor(id, levels=lvls)),
    mutate(V.TO, id=factor(id, levels=lvls)), by = "id")


#Fill in missing gender and major values - ifelse() will convert factors to numerical values so convert to character
VERTEX$gender.from <- ifelse(is.na(VERTEX$gender.from) == TRUE, as.factor(as.character(VERTEX$gender.to)), as.factor(as.character(VERTEX$gender.from)))

VERTEX$major.from <- ifelse(is.na(VERTEX$major.from) == TRUE, as.factor(as.character(VERTEX$major.to)), as.factor(as.character(VERTEX$major.from)))

#Remove redundant gender and major variables
VERTEX <- select(VERTEX, id, gender.from, major.from)

#rename variables
names(VERTEX) <- c("id", "gender", "major")

#Remove all the repeats so that we just have a list of each student and their characteristics
VERTEX <- unique(VERTEX)

```

Now we have both a Vertex and Edge list it is time to plot our graph!

```{r}
#Load the igraph package

library(igraph)

#First we will make an object that contains the graph information using our two dataframes EDGE and VERTEX. Notice that we have made "directed = TRUE" - our graph is directed since comments are being given from one student to another.

g <- graph.data.frame(EDGE, directed=TRUE, vertices=VERTEX)

#Now we can plot our graph using the force directed graphing technique - our old friend Fruchertman-Reingold!

plot(g,layout=layout.fruchterman.reingold)

#There are many ways to change the attributes of the graph to represent different characteristics of the newtork. For example, we can color the nodes according to gender.

plot(g,layout=layout.fruchterman.reingold, vertex.color=VERTEX$gender)

#We can change the thickness of the edge according to the number of times a particular student has sent another student a comment.

plot(g,layout=layout.fruchterman.reingold, vertex.color=VERTEX$gender, edge.width=EDGE$count)

```

## Part II

#In Part II your task is to [look up](http://igraph.org/r/) in the igraph documentation and modify the graph above so that:

#* Ensure that sizing allows for an unobstructed view of the network features (For example, the arrow size is smaller)
#* The vertices are colored according to major
#* The vertices are sized according to the number of comments they have recieved

```{r}
par(bg="pink")
#arrow
library(dplyr)
V(g)$vertex_degree <- degree(g)
#comment <- EDGE%>%group_by(to)%>%summarize(totalcomment=sum(count,na.rm=T))
#node.size<-setNames(comment$totalcomment,comment$to)
#vertex.size=node.size,
comment_count<-EDGE%>%group_by(to)%>%summarize(count=sum(count))
colnames(comment_count)<-c("id","count")
VERTEX <-left_join(VERTEX, comment_count, by = 'id')
VERTEX$count[is.na(VERTEX$count)] <- 0
plot(g,layout=layout.fruchterman.reingold,vertex.color=VERTEX$major,
     edge.arrow.size=.5,vertex.shape="circle",vertex.size=VERTEX$count*4,edge.curved=.1,vertex.frame.color="gray",vertex.label.cex=.8,)

```


## Part III

Now practice with data from our class. This data is real class data directly exported from Qualtrics and you will need to wrangle it into shape before you can work with it. Import it into R as a data frame and look at it carefully to identify problems.
```{r}
library(tidyverse)
library(readr)
library(igraph)
library(stringr)
#clean the data 
pn <- read.csv("hudk4050-classes.csv", skip=1, header = TRUE)
#recode name 
pn<-unite(pn,"name","First.Name","Last.Name",sep=" ")
#remove unpredictable characters from names 
pn$name<-str_replace(pn$name,"`","")
#remove unrelated columns
d_pn<-pn[-1,-8]
#create dataframe with two varialbe, students and name 
tidy_pn2<-d_pn%>%gather(key="class",value="course",Class.1:Class.6,-name, factor_key=TRUE,na.rm=TRUE)
#remove blank course
index2 <- which(tidy_pn2$course =="")
tidy_pn2<-tidy_pn2[-index2,] 
#remove duplicates (Danny!)
tidy_pn2<-unique(tidy_pn2)
#make all "course" letter capital
tidy_pn2$course=toupper(tidy_pn2$course)
#remove whitespace within "course"
tidy_pn2$course=gsub("[[:space:]]", "",tidy_pn2$course)
#recode wrong input of class
index3<-which(tidy_pn2$course=="41220")
tidy_pn2[index3,3]<-"4122"
#remove HUDK4050
index_remove4050<-which(tidy_pn2$course=="HUDK4050")
tidy_pn<-tidy_pn2[-index_remove4050,]
#change class type into factor instead of character
tidy_pn$course<-as.factor(tidy_pn$course)
#person-class matrix
persontoclass<-as.matrix(table(tidy_pn$name,tidy_pn$course)) 
persontoclass
#person-person matrix #这是还是不明白multiply怎么就能出现persontoperson
persontoperson=persontoclass%*%t(persontoclass)
persontoperson
#remove diagonal elements from person-person matrix
#igraph
g2 <-graph_from_adjacency_matrix(persontoperson,mode = c('undirected'),diag=F)
plot(g2,layout=layout.fruchterman.reingold)

#学习betweeness的内容完成下面部分
#**Who is the most central person in the network according to these two metrics? Write a sentence or two that describes your interpretation of these metrics**
  
#network metrics: Betweeness centrality and dregree centrality
sort(betweenness(g2),decreasing=T)
#the biggest betweenness centrality
g2between=betweenness(g2)
g2between[which(g2between==max(g2between))]

sort(degree(g2),decreasing=T)
#the biggest degree centrality
g2degree=degree(g2)
g2degree[which(g2degree==max(g2degree))]

#comment: Yifei Zhang (myself haha #_#) has the biggest betweenness centrality. Guoliang Xu, Hangshi Jin, Jiaao Qi, Jiacong Zhu, Jiahao Shen,Wenqi Gao, Xiyun Zhang and Yingxin Xie have the biggest degree centrality.
```

Please create a **person-network** with the data set hudk4050-classes.csv. To create this network you will need to create a person-class matrix using the tidyr functions and then create a person-person matrix using `t()`. You will then need to plot a matrix rather than a to/from data frame using igraph.

Once you have done this, also [look up](http://igraph.org/r/) how to generate the following network metrics:

* Betweeness centrality and dregree centrality. **Who is the most central person in the network according to these two metrics? Write a sentence or two that describes your interpretation of these metrics**

* Color the nodes according to interest. Are there any clusters of interest that correspond to clusters in the network? Write a sentence or two describing your interpetation.

```{r}
#remove first row
interest <- pn[-1,]$Which.of.these.topics.is.most.interesting.to.you.
ColorList = c("yellow", "darkblue", "green", "pink", "red")
V(g2)$color = ColorList[interest]
plot(g2,layout=layout.fruchterman.reingold,vertex.size=10)
#comment: Yes. There are clusters corresponding to the cluster in the network. In my opinion, there are more students interested in social network analysis in one cluster and more students interested in predicting modelling in the other cluster.
```

### To Submit Your Assignment

Please submit your assignment by first "knitting" your RMarkdown document into an html file and then comit, push and pull request both the RMarkdown file and the html file.
