---
title: "Assignment 3: K Means Clustering"
---

In this assignment we will be applying the K-means clustering algorithm we looked at in class. At the following link you can find a description of K-means:

https://www.cs.uic.edu/~wilkinson/Applets/cluster.html


```{r}
library(tidyverse)
library(tidyr)
library(tidyselect)
library(dplyr)
library(cluster)
library(factoextra)
library(klaR)
library(igraph)
library(ggplot2)

# ARTEMAS WANG ASSIGNMENT 3 SUBMISSION 
```

# Part I
```{r}


K1 <- Class_Motivation
K2 <- K1[c(0,-1)] # select all columns but the first one (id variable)
K3 <- na.omit(K2)
K3 <- scale(K3)
str(K3)

set.seed(123)
kmeans(K3, 2)
fit <- kmeans(K3, 3) # kmeans(df, # of clusters)
fit # 9 available components, including cluster
fit$cluster

K4 <- data.frame(fit$cluster,K3)
names(K4) <- c('Cluster','1','2','3','4','5') # the names need to be numbers. 
K4

# names_to needs to be numbers to change to as.numeric.
K5 <- K4 %>% pivot_longer(-Cluster, names_to = 'Weeks', values_to = 'Motivation') 
K5 
#K5 <- gather(K4, 'Cluster','Week', 2:6) %>% group_by(Week, Cluster)

# Find the average score for motivation
K6 <- K5 %>% group_by(Cluster, Weeks) %>% summarize(avg_score = mean(Motivation))
K6

# Change weeks to numeric and cluster as factors. 
K6$Week1 <- as.numeric(K6$Weeks)
K6$Cluster <- as.factor(K6$Cluster)

# Visualize the Cluster in line and scatterplot
ggplot(K6, aes(K6$Week1, K6$avg_score)) +
  geom_line(aes(color = K6$Cluster))

# ggplot(K6, aes(K6$Week1, K6$avg_score)) +
#   geom_point(aes(color = K6$Cluster))

K7 <- length(K6$Cluster)
K7


```


Look at the number of people in each cluster, now repeat this process for 3 rather than 2 clusters. Which cluster grouping do you think is more informative? Write your answer below:

##Part II

```{r}

DF1 <- hudk4050_classes
head(DF1)
# Merge First.name and Last.name variables to create unique ID, puts two columns into 1 column
DF1 <- unite(DF1, Name, `Last Name`, `First Name`, sep = " ", remove = TRUE)

# Reshape data to create a "course" variable (you will get a warning because there are missing cells)
DF2 <- gather(DF1, course.label, course, `Class 1`, `Class 2`, `Class 3`, `Class 4`, `Class 5`, 'Class 6', 
              na.rm = TRUE, convert = FALSE)

# Remove rows indicating HUDK4050 because all students are in this course and it will overwhelm the graph
DF2 <- filter(DF2, course > 0, course != "HUDK4050")

# Remove the "course.label" variable
DF2 <- dplyr::select(DF2, Name, course)

# Add a variable to be used to count the courses
DF2$Count <- 1

# Reshape the data to create a person x class matrix
DF3 <- spread(DF2, course, Count)

# This was a bit of a trick, for the matrix command to work the row names 
# needed to changed from an indice (1,2,3,etc) to the student names 
row.names(DF3) <- DF3$Name
DF3$Name <- NULL
DF3 <- ifelse(is.na(DF3), 0, 1) 

set.seed(100)
part2 <- kmodes(DF3, 3) # kmeans(df, # of clusters)
part2 # 9 available components, including cluster
part2$cluster

#Convert the data from data frame format to matrix format so it can be transposed
DF4 <- as.matrix(DF3)

#Transpose matrix to produce a person x person matrix
DF5 <- DF4 %*% t(DF4)

# set.seed(100)
# kmeans(DF5, 4)
# part2 <- kmeans(DF5, 4) # kmeans(df, # of clusters)
# part2 # 9 available components, including cluster
# part2$cluster
# DF5

g <- graph.adjacency(DF5, mode= "undirected")

plot(g, layout=layout.fruchterman.reingold, vertex.size=3, vertex.color=part2$cluster, margin = -0.4, edge.arrow.size = 0.4)

```


Using the data collected for Assignment 2 (which classes students were in), cluster the students, then redraw the graph of the class but color the students according the cluster they are in.

##Part III

```{r}

library(ggplot2)
survey <- HUDK405018_cluster
colnames(survey) #ResponseID, Q1_1
S1 <- survey %>% dplyr::select(9,18) 

addingrows <- data.frame("ResponseId" = 48:52, "Q1_1" = c('Beijing','Shanghai','Beijing','Shanghai','Beijing'))
addingrows
S2 <- rbind(S1, addingrows)

as.factor(S2$Q1_1)
as.factor(S2$ResponseId)

set.seed(100)
survey2 <- kmodes(S2, 3) # kmeans(df, # of clusters)
survey2$cluster
survey2
length(part2$cluster) # 52 people in class
length(survey2$cluster) # 47/52 people in class

ggplot(S2) +
  geom_jitter(aes(survey2$cluster, part2$cluster), color = part2$cluster)


```


In class activity 6 you clustered students in the class by the answers to a questionaire. Create a visualization that shows the overlap between these clusters and the clusters generated in part II.

## Please render your code as an .html file using knitr and Pull Resquest both your .Rmd file and .html files to the Assignment 3 repository.

