---
title: "class data"
author: "jiaxili"
date: "October 20, 2016"
---

```{r}
###install.packages("ggplot2") 
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r}
### upload the file "cluster-class-data.csv" Repository as a data frame called "K1"
K1 <- read.table("cluster-class-data.csv",header = TRUE, sep = ",")
```

### First part: cluster with regions (K group of sets)

```{r}
### Since algorithm will treat each row as a value belonging to a person, so remove the name column and leave the "regions" columns
K2 <- dplyr::select(K1, 14:16)
```

```{r}
### omit NA values
K3 <- na.omit(K2)
```

### Convert all regions into two new columns"Longtitude" and "Latitute"

```{r}
### Store longtitude and latitude into new varibles called "Lo" and "La"
Lo <-c(37.5665,14.0583,31.1343,22.5431,36.6512,33.0245,39.1368,35.0887,24.4798,30.4383,29.8683,22.5431,22.5431,28.6139,40.7128,23.0207,21.3069,39.9042,41.1415,39.9612,40.8923,41.4343)

La <- c(126.9780,108.2772,121.2829,114.0579,117.1200,71.5523,77.7147,92.4421,18.0894,84.2807,121.5440,114.0579,114.0579,77.2090,74.0059,113.7518,157.8583,116.4074,73.3579,82.9988,74.4774,74.0354)

### Create new table containg longtitude and latitude
K4 <- data.frame(Lo,La)
```

```{r}
### standardize the values so that they have the same range
K4 <- scale(K4)
```

```{r}
### kmeans with k3 dataset, and automatically have 2 cluster groups
fit <- kmeans(K4, 2) 
```

```{r}
### See how each observations are divided into 2 clusters
fit$cluster
```

```{r}
#Attach new column"cluster" to te original dataframe 
K4 <- data.frame(K4, fit$cluster)
```

```{r}
#Change the names of the variables with coordinates x & y 
names(K4) <- c("x", "Y", "cluster")
```

### Second part: cluster with questions other than regions (T group of sets)
```{r}
###### Since algorithm will treat each row as a value belonging to a person, so remove the name column and remove the "regions" columns
T2 <- dplyr::select(K1, 6:13)
###going to add on "cat" and "gif" columns later 
```

```{r}
###rename T2
names(T2) <- c("months"," siblings","times_sports","miles","friends_Andorid","movies","classes","states_visited")
```

```{r}
###select question3 and transfer string into numeric answers
T3<- dplyr::select(K1, QID3)
T3$QID3<-ifelse(T3$QID3 == 'No', 0, 1)
names(T3) <- "cat"
```

```{r}
###select question4 and transfer string into numeric answers
T4<- dplyr::select(K1, QID4)
T4$QID4<-ifelse(T4$QID4 == 'g-iff', 0, 1)
names(T4) <- "gif"
```

```{r}
###combine question3 and question4 into T2 original table
T2 <- data.frame(T2, T3,T4)
```

```{r}
### standardized for kmeans
T2 <- scale(T2)
```

```{r}
### compute kmeans with 2 clusters
fit2 <- kmeans(T2, 2) 
```

```{r}
### see what is in cluster
fit2$cluster
```

```{r}
### Add cluster to original table
T2 <- data.frame(T2, fit2$cluster)
```

```{r}
### rename T2 with new cluster column
names(T2) <- c("months"," siblings","times_sports","miles","friends_Andorid","movies","classes","states_visited","cat","git","cluster")
```

###Third part : now there are two groups of data: one group is  the "regions" data (K4), and the other is other "questions" (T2), the followings are trying to create visualization for those two groups of data

```{r}
###convert cluster from an "integer" format to a "factor"
###this step is unnessary for completing this assignment
K4$x <- as.numeric(K4$x)
K4$Y <- as.numeric(K4$Y)
K4$cluster <- as.factor(K4$cluster)

T2$months <- as.numeric(T2$months)
T2$times_sports <- as.numeric(T2$times_sports)
T2$miles <- as.numeric(T2$miles)
T2$friends_Andorid <- as.numeric(T2$friends_Andorid)
T2$movies <- as.numeric(T2$movies)
T2$classes <- as.numeric(T2$classes)
T2$states_visited <- as.numeric(T2$states_visited)
T2$cat <- as.numeric(T2$cat)
T2$git <- as.numeric(T2$git )
T2$` siblings` <- as.numeric (T2$` siblings`)
T2$cluster <- as.factor(T2$cluster)
```

```{r}
###some interesting plots for T2 & K4 seperately
 ggplot(T2, aes(months,times_sports, colour = cluster)) +geom_dotplot(stackgroups = TRUE, dotsize=0.5, stackdir = "up", stackratio = 1,binwidth = 0.5, binpositions="all")

ggplot(K4, aes(x, Y, colour = cluster)) + geom_line() + xlab("Longtitude") + ylab("Latitude")
```

```{r}
### Visualization with two groups of clusters
table1 <- table(K4$cluster,T2$cluster)
plot(table1, col = "dark blue", main = " Clusters Analysis",xlab = "Regions.Cluster", ylab= "Questions.Cluster")

###Better visualization with ggplot
###geom_raster
df1 <- data.frame(table1)
ggplot(df1, aes(df1$Var1,df1$Var2, colour = Freq)) +geom_raster(aes(fill = Freq)) +xlab ( "Regions.Cluster") + ylab ( "Questions.Cluster")

###geom_tile
ggplot(df1, aes(df1$Var1,df1$Var2, colour = Freq)) +  geom_tile(aes(fill = Freq)) +xlab ( "Regions.Cluster") + ylab ( "Questions.Cluster")
```

```{r}
###Interpretation with results

###To be conluded, people in Regions-cluster 1 (Asians) also group with Questions-cluster 1 while Geo-cluster 2 are spread between Question-clusters 1 & 2.

###In other words, students' clusters really depend on their different characteristics: where they grow, and where they live. For exmaple, those types of questions really group asians together (how long they stay in United States i.e.), and thus reflected in their regions.

###That is, if the Question-clusters 1& Regions-cluster 1 with freq in 8 can be considered as the most typical  asian-type students, then probably, Question-clusters 2 & Geo-cluster 1 are those who are not that much typical asians in behaviour.

###Lastly, there are 2 people are geographical remotes and also with distinctive answers for questions, so that they belong to Questions-cluster 2 & Regions-cluster 2.
