---
title: "Assignment 3: K Means Clustering"
#Allison Teevan
---

In this assignment we will be applying the K-means clustering algorithm we looked at in class. At the following link you can find a description of K-means:

https://www.cs.uic.edu/~wilkinson/Applets/cluster.html

Now, upload the file "Class_Motivation.csv" from the Assignment 3 Repository as a data frame called "K1""
```{r}

K1 <- read.csv("Class_Motivation.csv")

```

This file contains the self-reported motivation scores for a class over five weeks. We are going to look for patterns in motivation over this time and sort people into clusters based on those patterns.

But before we do that, we will need to manipulate the data frame into a structure that can be analyzed by our clustering algorithm.

The algorithm will treat each row as a value belonging to a person, so we need to remove the id variable.

```{r}
library(tidyverse)

#remove ID variable 
K2 <- K1 %>% select(2:6)

```

It is important to think about the meaning of missing values when clustering. We could treat them as having meaning or we could remove those people who have them. Neither option is ideal. What problems do you foresee if we recode or remove these values? Write your answers below:

#Removing people with missing values removes other valid data that we could otherwise include. If we treat the missing values as having meaning (ex. assign them the average score), we are assigning value to something that we do not know the value of.

We will remove people with missing values for this assignment, but keep in mind the issues that you have identified.


```{r}

K3 <- na.omit(K2) #This command create a data frame with only those people with no missing values. It "omits" all rows with missing values, also known as a "listwise deletion". EG - It runs down the list deleting rows as it goes.

```

Another pre-processing step used in K-means is to standardize the values so that they have the same range. We do this because we want to treat each week as equally important - if we do not standardise then the week with the largest range will have the greatest impact on which clusters are formed. We standardise the values by using the "scale()" command.

```{r}
#standardize the values
K3 <- scale(K3)

```


Now we will run the K-means clustering algorithm we talked about in class. 
1) The algorithm starts by randomly choosing some starting values 
2) Associates all observations near to those values with them
3) Calculates the mean of those clusters of values
4) Selects the observation closest to the mean of the cluster
5) Re-associates all observations closest to this observation
6) Continues this process until the clusters are no longer changing

Notice that in this case we have 5 variables and in class we only had 2. It is impossible to vizualise this process with 5 variables.

Also, we need to choose the number of clusters we think are in the data. We will start with 2.

```{r}

fit <- kmeans(K3,2)

#We have created an object called "fit" that contains all the details of our clustering including which observations belong to each cluster.

#We can access the list of clusters by typing "fit$cluster", the top row corresponds to the original order the rows were in. Notice we have deleted some rows.

fit$cluster

#We can also attach these clusters to the original dataframe by using the "data.frame" command to create a new data frame called K4.

K4 <- data.frame(K3, fit$cluster)

#Have a look at the K4 dataframe. Let's change the names of the variables to make it more convenient with the names() command.
names(K4) <- c("1", "2", "3", "4", "5", "cluster") 

```

Now we need to visualize the clusters we have created. To do so we want to play with the structure of our data. What would be most useful would be if we could visualize average motivation by cluster, by week. To do this we will need to convert our data from wide to long format. Remember your old friends tidyr and dplyr!

First lets use tidyr to convert from wide to long format.
```{r}

K5 <- K4 %>% gather("week","motivation",1:5)

```

Now lets use dplyr to average our motivation values by week and by cluster.

```{r}

K6 <- K5 %>% group_by(week, cluster) %>% summarize(avg = mean(motivation))

```

Now it's time to do some visualization: 

https://www.cs.uic.edu/~wilkinson/TheGrammarOfGraphics/GOG.html

And you can see the range of available graphics in ggplot here:

http://ggplot2.tidyverse.org/reference/index.html

We are going to create a line plot similar to the one created in the school dropout paper we looked at in class (Bowers, 2010). It will have motivation on the Y-axis and weeks on the X-axis. To do this we will want our weeks variables to be treated as a number, but because it was created from a variable name it is currently being treated as a character variable. You can see this if you click on the arrow on the left of K6 in the Data pane. Week is designated by "chr". To convert it to numeric, we use the as.numeric command. 

Likewise, since "cluster" is not numeric but rather a categorical label we want to convert it from an "integer" format to a "factor" format so that ggplot does not treat it as a number. We can do this with the as.factor() command.

```{r}
library(ggplot2)

K6$week <- as.numeric(K6$week)

K6$cluster <- as.character(K6$cluster)

```

Now we can plot our line plot using the ggplot command, "ggplot()".

- The first argument in a ggplot is the dataframe we are using: K6
- Next is what is called an aesthetic (aes), the aesthetic tells ggplot which variables to use and how to use them. Here we are using the variables "week" and "avg" on the x and y axes and we are going to color these variables using the "cluster" variable
- Then we are going to tell ggplot which type of plot we want to use by specifiying a "geom()", in this case a line plot: geom_line()
- Finally we are going to clean up our axes labels: xlab("Week") & ylab("Average Motivation")

```{r}

ggplot(K6, aes(week, avg, color = cluster)) + geom_line() + xlab("Week") + ylab("Average Motivation")

```

What patterns do you see in the plot?
#the plots are W and M shaped. Motivation varies up and down week by week.

It would be useful to determine how many people are in each cluster. We can do this easily with dplyr.

```{r}
library(dplyr)
K7 <- count(K4, cluster)
```

Look at the number of people in each cluster, now repeat this process for 3 rather than 2 clusters. Which cluster grouping do you think is more informative? Write your answer below:
#I think 3 clusters is more informative because it shows a third trend that is missing from the 2 cluster graph. Two of the cluster trends differ significantly in week 4 and 5.

```{r}

fit <- kmeans(K3,3)
K4 <- data.frame(K3, fit$cluster)
names(K4) <- c("1", "2", "3", "4", "5", "cluster") 
K5 <- K4 %>% gather("week","motivation",1:5)
K6 <- K5 %>% group_by(week, cluster) %>% summarize(avg = mean(motivation))
K6$week <- as.numeric(K6$week)
K6$cluster <- as.character(K6$cluster)

#plot using ggplot
ggplot(K6, aes(week, avg, color = cluster)) + geom_line() + xlab("Week") + ylab("Average Motivation")

#number of people in each cluster
(K7 <- count(K4, cluster))

```

Once you have done this, save both of your plots to the Assignment 5 file.

##########################################################################################
##Part II ################################################################################
##########################################################################################

Using the data collected for Assignment 2 (which classes students were in), cluster the students, then redraw the graph of the class but color the students according to the 3cluster they are in.

```{r}
library(tidyverse)
library(ggplot2)

rm(list = ls())
data <- read.csv("hudk4050-classes.csv")

#combine name to be full 
data <- unite(data, "Name", c("First.Name", "Last.Name"), sep = "_")

#remove spaces in names column 
data$Name <- (gsub(" ", "", data$Name))
#all uppercase
data$Name <- toupper(data$Name)

data2 <- gather(data, class_type, course_name, 2:7)

#remove class_type and ID 
data2$class_type <- NULL

#remove spaces from course names 
data2$course_name <- (gsub(" ", "", data2$course_name))

#replace tricky entries
data2$course_name <- (gsub("HUDK4050", "4050", data2$course_name))
data2$course_name <- (gsub("4050", "HUDK4050", data2$course_name))
data2$course_name <- (gsub("HUDMHUDK4050", "HUDM4050", data2$course_name))

data2$course_name <- (gsub("HUDM4125", "4125", data2$course_name))
data2$course_name <- (gsub("4125", "HUDM4125", data2$course_name))

data2$course_name <- (gsub("HUDM5026", "5026", data2$course_name))
data2$course_name <- (gsub("5026", "HUDM5026", data2$course_name))

data2$course_name <- (gsub("HUDM5126", "5126", data2$course_name))
data2$course_name <- (gsub("5126", "HUDM5126", data2$course_name))

#remove duplicates
data2 <- unique(data2)

#remove HUDK4050 class for everyone
data2 <-data2[(data2$course_name!="HUDK4050"),]

#add column of 1s
data2$count <- "1"

#create new matrix how we want it
data3 <- spread(data2, course_name, count)

#add ID number to name list so we can match them later 
data3$id2 <- seq(1,53,1)

Final2 <- select(data3, 1, 61)

#remove column of V1 (blank) and ID
data3$V1 <- NULL
data3$id2 <- NULL

#replace NA with zero
data3[is.na(data3)] <- 0
#move full name to be row name
row.names(data3) <- data3$Name
data3$Name <- NULL

```

#create the clusters 
```{r}
fit2 <- kmeans(data3, 3)
#I think there should be three clusters- those in Learning Analytics program, the Applied Stats program, and everyone else. People in the same program will be taking similar classes. 

K4 <- data.frame(data3, fit2$cluster)

#number of people in each cluster
count(K4, fit2.cluster)

Final2 <- data.frame(Final2, fit2$cluster)

```


```{r}
#trying to multiply matrices to get matrix to plot 
data3 <- as.data.frame(sapply(data3, as.numeric))
data3 <- as.matrix((data3))
data4 <- t(data3)

data5 <- data3 %*% data4

#add back names
row.names(data5) <- Final2$Name
colnames(data5) <- Final2$Name

library(igraph)

graph3 <- graph_from_adjacency_matrix(data5, weighted = TRUE,
  diag = FALSE, add.colnames = NULL, add.rownames = NA)

#plot graph
plot(graph3,layout=layout.fruchterman.reingold,edge.arrow.size=0.01, vertex.color=K4$fit2.cluster)

```

##Part III

In class activity 6 you clustered students in the class by the answers to a questionaire. Create a visualization that shows the overlap between these clusters and the clusters generated in part II.

```{r}
#READ IN DATA FROM CLASS ACTIVITY 6

#Load data
DF1 <- read.csv("HUDK405019-clustering.csv", header = TRUE)

#name manipulation
DF1 <- unite(DF1, "Name", c("First.Name", "Last.Name"), sep = "_")
#remove spaces in names column 
DF1$Name <- (gsub(" ", "", DF1$Name))
DF1$Name <- toupper(DF1$Name)

#add ID for later
DF1$id1 <- seq(1,50,1)

Final1 <- select(DF1, 1, 16)

row.names(DF1) <- DF1$Name
DF1$Name <- NULL

#Remove location variables
DF2 <- select(DF1, 1:11)

#Remove any characters
DF2 <- DF2 %>% mutate_all(funs(gsub("[a-zA-Z]", "", .)))

#Convert all variables to numeric
DF2 <- DF2 %>% mutate_all(funs(as.numeric(.)))

#Scale the data so that no variable has undue influence
DF2 <- as.data.frame(scale(DF2))
 
#Replace missing values with average score EG - zero
DF2 <- DF2 %>% mutate_all(funs(ifelse(is.na(.) == TRUE, 0, .)))

DF3 <- select(DF1, 13:14)
#Change names for convenience
names(DF3) <- c("lattitude", "longitude")

#Remove any characters and common punctuation
DF3 <- DF3 %>% mutate_all(funs(gsub("[a-zA-Z]", "", .)))
DF3 <- DF3 %>% mutate_all(funs(sub("[?]", "", .)))

#Remove anything after the first non-numeric character in lattitude
DF3$lattitude <- sub(",.*$","", DF3$lattitude) 
DF3$lattitude <- sub("°.*$","", DF3$lattitude)

#Remove anything before the first non-numeric character in longitude
DF3$longitude <- gsub(".*,","",DF3$longitude)
DF3$longitude <- sub("°.*$","", DF3$longitude)

#Convert all variables to numeric
DF3 <- DF3 %>% mutate_all(funs(as.numeric(.)))

#kmeans
fit1 <- kmeans(DF2, 3) 
Final1 <- data.frame(Final1, fit1$cluster)

```


```{r}
#Note: fit1 is from Activity 6 and fit2 is from Assignment 2

Final <- left_join(Final1,Final2, by = c("Name")) 

#remove un-needed rows
Final <- select(Final,Name,Cluster1=fit1.cluster,Cluster2 = fit2.cluster)

#remove students who didn't answer both surveys
Final <- Final[!(is.na(Final$Cluster1)),]
Final <- Final[!(is.na(Final$Cluster2)),]
```


```{r}
#install.packages("vcd")
library(vcd)

#create matrix counting number of students in each group pairing
P1 <- structable(Final$Cluster1 ~ Final$Cluster2)
P2 <- as.matrix(P1)

#create heatmap
#This shows how much overlap there are between the groups of clusters
heatmap(P2)

```

## Please render your code as an .html file using knitr and Pull Request both your .Rmd file and .html files to the Assignment 3 repository.

