---
title: 'HUDK4050: Class Activity 6'
author: "Charles Lang"
date: "10/23/2018"
output: html_document
---
# Data Management
```{r}
#Load data
DF1 <- read.csv("HUDK405018-cluster.csv", header = TRUE)
  
#Wrangle data
DF1 <- select(DF1,Q1_1,Q1_2) 

```

# Find lattitudes & longitudes for cities
```{r}
install.packages("ggmap")
library(ggmap)

#Request lattitude and longitude from Google Maps API
DF2 <- geocode(as.character(DF1$Q1_1), output = "latlon", source = "dsk")
DF2<-DF2[-c(1,2),]
```

Now we will run the K-means clustering algorithm we talked about in class. 
1) The algorithm starts by randomly choosing some starting values 
2) Associates all observations near to those values with them
3) Calculates the mean of those clusters of values
4) Selects the observation closest to the mean of the cluster
5) Re-associates all observations closest to this observation
6) Continues this process until the clusters are no longer changing

Notice that in this case we have 5 variables and in class we only had 2. It is impossible to vizualise this process with 5 variables.

Also, we need to choose the number of clusters we think are in the data. We will start with 4.

```{r}

fit_map <- kmeans(DF2, 2) 

#We have created an object called "fit" that contains all the details of our clustering including which observations belong to each cluster.

#We can access the list of clusters by typing "fit$cluster", the top row corresponds to the original order the rows were in. Notice we have deleted some rows.

fit_map$cluster

#We can also attach these clusters to te original dataframe by using the "data.frame" command to create a new data frame called K4.

DF3 <- data.frame(DF2, fit_map$cluster)

#Have a look at the K4 dataframe. Lets change the names of the variables to make it more convenient with the names() command.

names(DF3) <- c("lon", "lat", "cluster") #c() stands for concatonate and it creates a vector of anything, in this case a vector of names.

```

# Visualize your clusters in ggplot
```{r}
#Create a scatterplot that plots location of each student and colors the points according to their cluster 

g<-ggplot(data=DF3, aes(x=lon,y=lat, xlab = "longitude", ylab="latitude", colour=DF3$cluster)) + geom_point()
g

```

# Can you group students according to their weekly activities?
How much time did you spend this week in hours: - Coding (R, Java, SPSS, etc)
How much time did you spend this week in hours: - Writing papers for class
How much time did you spend this week in hours: - Reading/videos for class
How much time did you spend this week in hours: - Watching videos for fun
How much time did you spend this week in hours: - Cooking
How much time did you spend this week in hours: - Using social media
How much time did you spend this week in hours: - Social activities (movies, restaurants, etc)
```{r}
#Load data
DF1 <- read.csv("HUDK405018-cluster.csv", header = TRUE, na.strings=c("","NA"), stringsAsFactors = FALSE)

#Wrangle data
DF1 <- select(DF1,Q2_1:Q2_7)
DF2<-DF1[-c(1,2),]
names(DF2) <- c("1", "2", "3","4","5","6", "7")

# turn NAs into zeroes
DF2[is.na(DF2)] <- 0
DF3<-DF2
#I know there's probably a faster way to do this but I got tired of searching for it
DF3$'1' <-as.numeric(DF2$'1')
DF3$'2' <-as.numeric(DF2$'2')
DF3$'3'<-as.numeric(DF2$'3')
DF3$'4'<-as.numeric(DF2$'4')
DF3$'5'<-as.numeric(DF2$'5')
DF3$'6'<-as.numeric(DF2$'6')
DF3$'7'<-as.numeric(DF2$'7')

#Rectifying the earlier stupid mistake to change names to numbers
names(DF3)<-c("coding","writing papers","read/vid", "watch videos","cooking","social media", "going out")

##maybe don't use this. scales down my hours way low
#DF3<-scale(DF2)

#Kmeans for activity of students
activity <- kmeans(DF3,3)
activity$cluster

DF4<-data.frame(DF3)
DF4$cluster<-activity$cluster

DF5<-gather(DF4,activity,hours,1:7,factor_key=TRUE)

DF6 <- DF5%>% group_by(activity, cluster)%>% dplyr::summarize(avg=mean(hours, na.rm=TRUE))

DF6$activity <- as.numeric(DF6$activity)

DF6$cluster <- as.factor(DF6$cluster)

ggplot(DF6,aes(activity,avg, colour=DF6$cluster))+geom_line()+labs(x="activity:1=coding, 2=writing, 3=read/vid, 4=vid/fun, 5=cooking, 6=social media, 7=activities", y="Average hours")

DF7 <- count(DF5, cluster, sort=FALSE)
```

